---
title: "BMIN 7054 Homework 5"
author: "Mike Lape"
date: "March 8, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, get_dat, warning=FALSE, message=FALSE}
library(pROC)
data(aSAH)
dat <- aSAH
```

&nbsp;


1.	Write a short note on the protein biomarker ndka from internet sources (Max 5 sentences).  
&nbsp;
NDKA better known as NME1, nucleoside diphosphate kinase A, is an enzyme
encoded by the NME1 gene.  It is found in most tissues and is repsonsible 
for moving the terminal phosphate amongst various nucleoside diphosphates.
It is also believed to be involved in the "ischemic cascade after stroke".  It
is considered an early biomarker for a stroke.  It has also been looked at 
as a biomarker for color rectal cancer.

* https://en.wikipedia.org/wiki/NME1  
* Otero-estévez O, De chiara L, Barcia-castro L, et al. Evaluation of serum nucleoside diphosphate kinase A for the detection of colorectal cancer. Sci Rep. 2016;6:26703.  
* Han X, Gao Y, Ma B, et al. The Clinical Relevance of Serum NDKA, NMDA, PARK7, and UFDP Levels with Phlegm-Heat Syndrome and Treatment Efficacy Evaluation of Traditional Chinese Medicine in Acute Ischemic Stroke. Evid Based Complement Alternat Med. 2015;2015:270498.  


&nbsp;

2.   Fit logistic regression model with response var 'outcome' and predictors
s100b, log(ndka), gender, and age.  Identify significant predictors.  Identify
the most significant predictor.  Check goodness-of-fit of the model.  Write the
prediction algorith.  
```{r, fig.width=8, fig.height=8}
# log transform ndka
dat$log_nkda <- log(dat$ndka)

# Create the model
mod <- glm(outcome ~ s100b + log_nkda + gender + age, family = binomial, data = dat)

# Examine model
summary(mod)

# The following coefficients are calculated to be significant:
# Coef            Value           P-Val
# Intercept        -5.45          0.0004
# s100b             5.28          7.94e-05
# log_nkda          0.93          0.0197 
# genderFemale     -1.17          0.0235
# age               0.04          0.0477

# All predictors are significant but the most significant are s100b
# and log_ndka.  Leaving out the intercept.

# Goodness of fit:
# Residual deviance (output by glm): 110.34
# Degrees of Freedom (output by glm): 108

# Calculate p-val for fit, pchisq(residual dev, deg of free, lower.tail = F)
pval <- pchisq(110.34, 108, lower.tail = FALSE)
pval
# The calculated p-value is 0.42, meaning this model is a good fit.

```

Z = -5.45 + 5.28 * s100b + 0.93 * log_nkda - 1.17 * genderFemale + 
             0.04 * age 


Pr(Poor Outcome) = e^(Z) / ( 1 + e^(Z))

Pr(Good Outcome) = 1 / (1 + e^(Z))

&nbsp;

3. Obtain the confusion matrix of the model in question 2.  Find the accuracy
of the prediction model.  
```{r, fig.width=8, fig.height=8}
# To generate confusion matrix we need to predict some data, so let's 
# predict the training data.

# Running the training data through the model to get some predictions
pred <- predict.glm(mod, newdata = dat, type = "response" )
pred_class <- ifelse(pred >= 0.5, "Poor", "Good")

conf <- table(dat$outcome, pred_class)
rownames(conf) = c("Obs Good", "Obs Poor")
colnames(conf) = c("Pred Good", "Pred Poor")
conf

miss <- ((conf[2] + conf[3]) / nrow(dat)) * 100

paste("The misclassification rate is: ", round(miss, 2), "%")
paste("The accuracy is: ", round((100 - miss) , 2), "%")
```
&nbsp;


4.  Write formally the link function of the model, which we will use as
biomarker.  

Biomarker_X = Link Function = -5.46 + 5.28 * s100b + 0.93 * log_nkda 
                                - 1.17 * genderFemale + 0.04 * age

&nbsp;

5.  Get the non-parametric density curves of biomarker X for each outcome
(Good, Poor), and comment on the densities.  
```{r, fig.width=8, fig.height=8}
# Generate the biomarker X values
link <- predict.glm(mod, newdata = dat, type = "link" )

# Subset dat based on outcome.
good <- subset(link, dat$outcome == "Good")
poor <- subset(link, dat$outcome == "Poor")

# Now plot the density curves.
dens_good <- density(good)
dens_poor <- density(poor)

# Good density curve
plot(dens_good, col = "orange", lwd = 4, xlim = c(-4, 13), 
     xlab = "Biomarker X", ylab = "Density", 
     main = "Non-parametric Density Curves" )

# Poor density curve
lines(dens_poor, col = "blue", lwd = 4)

# Add ticks for data according to outcomes
rug(good, side = 1, col = "orange")
rug(poor, side = 3, col = "blue")

legend("topright", legend = c("Good", "Poor"), pch = c(16,16), 
       col = c("orange", "blue"))

# The density plot for our biomarker X values looks pretty good.  Poor outcomes
# tend to have larger values for biomarker X whereas good outcomes tend to
# have smaller (more negative) biomarker X values.  Also, while there is some overlap
# of poor and good outcomes in the range of -1 to 0 for biomarker X values, the good outcomes
# in this region far outnumber the poor outcomes.

```

&nbsp;


6.  Outline how a diagnostic test would look for biomarker X.   
Test is positive for 'poor' prognosis if biomarker_X >= cutoff.    
Test is negative for 'good' prognosis if biomarker_X < cutoff.  

We need to generate our ROC curve to get the optimal value of cutoff which
we will do below.




&nbsp;

7.  Invoke the 'roc' function with link function X as a biomarker.  Plot the ROC
curve.  
```{r, fig.width=8, fig.height=8}
df <- data.frame(link = c(good, poor), outcome = rep(c("Good","Poor"), c(72,41)))

# Poor = 1
# Good = 0 
# so direction is <
my_roc <- roc(df$outcome, df$link, levels = c("Good", "Poor"), direction = "<")

# Stuff the sensitivity and specificities into a dataframe
roc_out <- data.frame(sens = my_roc$sensitivities, 
                      spec = my_roc$specificities, cuts = my_roc$thresholds)
plot.roc(my_roc)
```

&nbsp;


8.  Find the 95% confidence interval for AUC.  Comment on the utility of 
biomarker X.
```{r}

ci <- ci.auc(my_roc)
paste(round(ci[1], 3), '<=', round(ci[2], 3), "<=", round(ci[3], 3))
# The null hypothesis that AUC = 0.5 is well outside the lower bound
# of the 95% confidence interval, indicating that this model is significantly 
# better than 50-50 chance.  Therefore biomarker X is useful and gives 
# decent sensitivity and specificity and would make a good diagnostic test.
```

9. Develop a diagnostic test based on biomarker X.
```{r}
# To develop a diagnostic test we need to decide on the best cutoff.
roc_out$dist <- (roc_out$sens -1)^2 + (roc_out$spec - 1)^2 

# Our best cutoff is the one wiht the minimal square distance.
roc_out[which(roc_out$dist == min(roc_out$dist)),]
```
Test is positive for 'poor' prognosis if biomarker_X >= -0.889.      
Test is negative for 'good' prognosis if biomarker_X < -0.889.   

10. Compare the biomarkes s100b and X.  
To compare two different biomarkers for which we have ROC curves, we  
simply need to compare the AUC for each.  The one with the larger AUC is 
the better performing biomarker.  In this case the AUC for biomarker X is 0.824 
and the AUC for s100b (as obtained from Dr. Rao's handout) is 0.74.  Thus
biomarker X is the better biomarker.