---
title: "BMIN 7054 Homework 4"
author: "Mike Lape"
date: "March 1, 2018"
output:
  pdf_document: default
---
&nbsp;

```{r, get_dat, warning=FALSE, message=FALSE}
library(pROC)
data(aSAH)
dat <- aSAH
```
&nbsp;


1.	Write a short note on the protein biomarker s100b from internet sources (Max 5 sentences)  
&nbsp;

S100B is a protein from the S100 family that binds calcium.  S100B is found in many different cell types, but it is found most in astroglial and Schwann cells.  It is found to reside in both the cytoplasm and the nucleus of cells.  Since s100b helps manage cellular calcium concentrations it can affect a large number of different biological processes including cell proliferation.  S100B can be used as a biomarker of brain damage because as brain cells are damaged or destroyed S100B flows into the bloodstream where an increased concentration can be detected.

Yardan T, Erenler AK, Baydin A, Aydin K, Cokluk C. Usefulness of S100B protein in neurological disorders. J Pak Med Assoc. 2011;61(3):276-81.  
https://en.wikipedia.org/wiki/S100B  

&nbsp;

2.   Cross-tabulate 'outcome' and 'gos6'. Comment on the table.
```{r, fig.width=8, fig.height=8}
table(dat$outcome, dat$gos6)
# There appears to be a qualitatively decent classification of outcome between
# good and poor based on the gos6 level. Specifically gos6 levels of 4 or 5,
# much more so 5 have good outcomes whereas low levels, 1 or 3 (oddly none of
# level 2) have poor outcomes.  This might mean that gos6 could be a decent
# biomarker indicating a particular probable outcome.

```

&nbsp;

3. Cross-tabulate 'outcome' and'wfns'. Comment on table.
```{r, fig.width=8, fig.height=8}
table(dat$outcome, dat$wfns)
# The level of wfns appears to have no correlation to the outcome a pateint
# has, meaning wfns would not make a good biomarker, because the level of
# wfns gives no information about the possible outcome.
```
&nbsp;


4.  Explore whether 'nkda' could be a good biomarker.  Because there is so much variation in nkda use the Ln transformation of the variable, (log(ndka)).  Segregate into 'good' or 'bad' outcomes.  Plot the density curves of log(ndka) for 'good' and 'bad' outcomes in the same graph.  Comment on the graph.
```{r, fig.width=8, fig.height=8}
# Log transform of nkda values
dat$log_nkda <- log(dat$ndka)
summary(dat$log_nkda)

# Subset dat based on outcome.
good <- subset(dat, dat$outcome == "Good")
poor <- subset(dat, dat$outcome == "Poor")

# Now plot the density curves.
dens_good <- density(good$log_nkda)
dens_poor <- density(poor$log_nkda)

# Good density curve
plot(dens_good, col = "orange", lwd = 4, xlim = c(1, 6.5), 
     xlab = "Possible Biomarker - Log(ndka)", ylab = "Density", 
     main = "Non-parametric Density Curves" )

# Poor density curve
lines(dens_poor, col = "blue", lwd = 4)

# Add ticks for data according to outcomes
rug(good$log_nkda, side = 1, col = "orange")
rug(poor$log_nkda, side = 3, col = "blue")

legend("topright", legend = c("Good", "Poor"), pch = c(16,16), 
       col = c("orange", "blue"))

# The log(ndka) correpsonding to a good outcome has a slightly lower
# mean than the mean for poor outcomes.  However, there is quite a bit 
# of overlap between the two densities, indiciating that there might 
# not be good enough separation between log(ndka) 
# values that are associated with good outcomes and those associated
# with poor outcomes.  So this calls into question whther log(ndka)
# could be a useful biomarker.  Based on the plot the best cutoff
# for log(ndka) to split the data into good or poor outcomes is around
# 2.25.

```

&nbsp;

5.  Invoke the 'roc' function with log(ndka) as a biomarker.  Plot the ROC curve that comes with it.
```{r, fig.width=8, fig.height=8}
my_roc <- roc(dat$outcome, dat$log_nkda, levels = c("Good", "Poor"), direction = "<")

# Stuff the sensitivity and specificities into a dataframe
df <- data.frame(sens = my_roc$sensitivities, spec = my_roc$specificities, cuts = my_roc$thresholds)
plot.roc(my_roc)

```

&nbsp;


6.  Find the 95% confidence interval for AUC.  Comment on the utility of the biomarker log(ndka).
```{r}
# Get the actual area under the curve, use a smoothed form of ROC instead of
# jagged version above.
smooth_roc <- roc(dat$outcome, dat$log_nkda, smooth = T, 
                  levels = c("Good", "Poor"), direction = "<")

plot.roc(smooth_roc)

ci <- ci.auc(smooth_roc)
paste(round(ci[1], 3), '<=', round(ci[2], 3), "<=", round(ci[3], 3))
# The null hypothesis that AUC = 0.5 is within the lower bound  
# of the 95% confidence interval, indicating that we cannot reject 
# the null hypothesis here so the log(ndka) biomarker is no better
# than 50-50 chance and thus is not useful, and definitely not worth
# the work of using it for a diagnostic test.


```

&nbsp;

7.  Develop a diagnostic test based on log(ndka).
```{r, fig.width=8, fig.height=8}
# To develop a diagnostic test we need to decide on the best cutoff.
df$dist <- (df$sens -1)^2 + (df$spec - 1)^2 

# Our best cutoff is the one wiht the minimal square distance.
df[which(df$dist == min(df$dist)),]
```
Test is positive for 'poor' prognosis if log(ndka) $>=$ 2.54.  
Test is negative for 'good' prognosis if log(ndka) < 2.54.  

&nbsp;


8.  Compare the biomarkers s100b and log(ndka).  
To compare two different biomarkers for which we have ROC curves, we  
simply need to compare the AUC for each.  The one with the larger AUC is 
the better performing biomarker.  In this case the AUC for log(ndka) is 0.60 
and the AUC for s100b (as obtained from Dr. Rao's handout) is 0.74.  Thus
s100b is a much better biomarker.