---
title: "BMIN 7054 Homework 1"
author: "Mike Lape"
date: "January 27, 2018"
output:
  word_document: default
---
&nbsp;

1.	What is the size of the data set and show first 6 rows.
```{r, fig.width=8, fig.height=8}

# Load in birthwt data
data("birthwt", package = 'MASS')

# The birthwt dataframe has 10 columns/variables, and 189 rows/observations.
dim(birthwt)


# The top 6 rows are as follows:
head(birthwt)

```

&nbsp;

2. Describe the data
```{r, fig.width=8, fig.height=8, eval=FALSE}
# Use the help page on this dataset to determine what each variable describes.
help(birthwt, package = MASS)
```
Low: Indicator of birth weight < 2.5 kg [Categorical - binary]  
age: Age of mother in years [Numerical]  
lwt: Mother's weight in pounds at last menstrual period [Numerical]  
race: Mother's race, 1 = white, 2 = black, 3 = other [Categorical]  
smoke: Smoking status during pregnancy [Categorical - binary]  
ptl: Number of previous premature labors [Numerical]  
ht: history of hypertension [Categorical - binary]  
ui: Presence of uterine irritability [Categorical - binary]  
ftv: Number of physician visits during 1st trimester [Numerical]  
bwt: Birth weight in grams [Numerical]  

&nbsp;

3. Determine datatype or class of each variable.
```{r, fig.width=8, fig.height=8}
# Remove bwt column.
birthwt1 <- birthwt[,-10]

# Generate list of types of each variable
sapply(birthwt1, class)

# Some of these variables are listed as integer but should changed to factors.

```

&nbsp;


4. Change race to factor and count number of each
```{r, fig.width=8, fig.height=8}

# Change race to factor.
birthwt1$race <- factor(birthwt1$race, labels=c("_White","_Black","_Other"))

# Display totals of each race in table
# 96 White, 26 Black, 67 Other
table(birthwt1$race)

```

&nbsp;


5. Postulate logistic regression with low as response variable.
```{r, fig.width=8, fig.height=8}

# Clean up the rest of the data first. 
birthwt1$low   <- factor(birthwt1$low, labels = c("Normal", "Low"))
birthwt1$smoke <- factor(birthwt1$smoke, labels = c("_No", "_Yes"))
birthwt1$ht    <- factor(birthwt1$ht, labels = c("_No", "_Yes"))
birthwt1$ui    <- factor(birthwt1$ui, labels = c("_No", "_Yes"))

# Postulate the model:
mod <- glm(low ~ ., data = birthwt1, family = binomial)
summary(mod)


```

&nbsp;



6. Fit the model to the data, write the prediction equation.  
Model was fitted in question 5 above.  The prediction equation can be found below.  

&nbsp;
Z = 0.4806 - 0.0295 * age - 0.0154 * lwt + 1.272 * race_Black + 0.880 * race_Other + 0.939 * smoke_Yes + 0.543 * ptl + 1.863 * ht_Yes + 0.768 ui_Yes + 0.065 * ftv

Pr(low birth weight) = e^(Z) / ( 1 + e^(Z))

&nbsp;

7. Identify the significant predictions, and interpret the coefficient for race  

The significant predictors, those with a p-val < 0.05 are  
lwt        (P-val = 0.02580)  
race_black (P-val = 0.01584)  
race_other (P-val = 0.04576)  
smoke_Yes  (P-val = 0.01957)  
ht_Yes     (P-val = 0.00756)  

The factor race is shown to have a significant impact on the prediction for low birth weight.
In this model race_white is held as the reference and p-values as well as coefficients are 
calculated for both race_black (p-val = 0.01584, coeff = 1.272 ) and race_other (p-val = 0.04576, 
coeff = 0.881).  The positive coefficients indicate that mothers of race black or other
have increased likelihood of giving birth to an underweight baby as compared to white mothers.


&nbsp;

8. Interpret the coefficients.   

The insignificant (by p-val) coefficients, age, ptl, ui, and ftv have the following coefficients, -0.0295, 0.543, 0.768, and 0.065.  Age has a slightly negative coefficient indicating that as a mother ages the chances of a low birth weight are less.  On the other hand, ptl and ui have positive coefficients, indicating that the more previous premature labors (ptl) or the presence of a uterine infection (ui) increase the odds of having a baby of low birth weight.  Finally, the more a mother goes to the doctor during the first trimester (ftv) actually increases the chances of a low birth weight child.
The rest of the coefficients are considered significant by the p-val.  As stated in question 7, the coefficients of race_black and race_other indicate that falling into either category increases the likelihood of a low weight birth.  The weight of the mother at the last menstrual period (lwt) has a barely negative coefficient (-0.015) indicating that as the mother's weight increases the chance of a low weight birth decreases.  On the other hand both smoke and ht have positive coefficients, 
0.939 and 1.863 respectively, indicating that both smoking and having hypertension increase
the odds of a low birth weight.


&nbsp;

9. Lay out the odds ratios along with their 95% confidence intervals. 
```{r, fig.width=8, fig.height=8}
# The odds ratios:
exp(mod$coefficients)

# The 95% CI:
exp(confint(mod))
```

&nbsp;

10. Work out the confusion matrix. Determine the misclassification rate. 
```{r, fig.width=8, fig.height=8, warning = FALSE, message = FALSE}
# Running the training data through the model to get some predictions
pred <- predict.glm(mod, type = "response" )

# Classify the prediction, and build confusion matrix. 
pred_class <- ifelse(pred >= 0.5, 1,0)
conf <- table(birthwt1$low, pred_class)
rownames(conf) = c('Obs_Normal', 'Obs_Low')
colnames(conf) = c('Pred_Normal', 'Pred_Low')
conf  

# Calculate misclassification rate using confusion matrix.
miss <- round(((conf[2] + conf[3]) / (conf[1] + conf[2] + conf[3] + conf[4])) * 100,2)


cat("The misclassification rate is: ", miss, "%")
```

&nbsp;

11. Summarize your findings in a few lines.   
The misclassification rate of this logistic regression model is 25.93% indicating that more than 
1 in every 4 prediction on average would be misclassified.  The test set used for the determination
of this misclassification rate is the same as the training set for the model, so with 
real test data the model would likely have a higher misclassification rate.

The model did output some important predictions, including that if you are black or other race when compared to white mother's you have an increased likelihood of a low birth weight.  Another important indicator is that if the mother has hypertension a low weight birth is 86.6% more likely.  This model seems to reinforce two pieces of advice often given to expecting mothers, namely do not smoke, and work to lower the blood pressure (hypertension).

&nbsp;